Supplement to ``{\mytitle}''

### 2023-08-31 si/si.Rnw
Made most of final changes, need to adjust how supplement and ms interact

%% copied from pnas-new.cls
\usepackage{amsmath,amsfonts,amssymb,graphicx}

%%%% parameters %%%%%%%%%%5
\newcommand\Wsat{W_{\mathrm{sat}}}
\newcommand\muIR{\mu_{IR}}
\newcommand\muEI{\mu_{EI}}
\newcommand\transmission{\beta}
\newcommand\seasAmplitude{a}
\newcommand\rainfallExponent{r}
\newcommand\muRS{\mu_{RS}}
\newcommand\vaccineEfficacy{\theta}
\newcommand\muBirth{\mu_S}
\newcommand\muDeath{\delta}
\newcommand\choleraDeath{\delta_{C}}
\newcommand\symptomFrac{f}
\newcommand\asymptomRelativeInfect{\epsilon}
\newcommand\asymptomRelativeShed{\epsilon_{W}}
\newcommand\Wbeta[1]{\beta_{W#1}}
\newcommand\Wremoval{\delta_W}
\newcommand\Wshed{\mu_W}
\newcommand\mixExponent{\nu}
\newcommand\sigmaProc{\sigma_{\mathrm{proc}}}
\newcommand\reportRate{\rho}
\newcommand\obsOverdispersion{\psi}
\newcommand\phaseParm{\phi}
\newcommand\transmissionTrend{\zeta}
\newcommand\Binit{\xi}
\newcommand\vaccClass{Z}
\newcommand\vaccCounter{z}
\newcommand\modelCounter{m}
\newcommand\Whur[1]{\beta_{W#1}^{hm}}
\newcommand\hHur[1]{h_{#1}^{hm}}
\newcommand\Iinit{I_{0,0}}
\newcommand\Einit{E_{0, 0}}
\newcommand\reportChange{\delta}

@@

%% EDITING MACROS
\usepackage{color}
\usepackage[normalem]{ulem}% to use \sout in feedback commands
\usepackage{bm}
% orange for EI
\definecolor{orange}{rgb}{1,0.5,0}
\newcommand\ei[2]{\sout{#1} \textcolor{orange}{#2}}
\newcommand\eic[1]{\textcolor{orange}{[#1]}}
% green for JW
\definecolor{green}{rgb}{0,0.5,0}
\newcommand\jw[2]{\sout{#1} \textcolor{green}{#2}}
\newcommand\jwc[1]{\textcolor{green}{[#1]}}
% purple for JJ
\definecolor{purple}{rgb}{0.5,0,1}
\newcommand\jj[2]{\sout{#1} \textcolor{purple}{#2}}
\newcommand\jjc[1]{\textcolor{purple}{[#1]}}
% cyan for AR
\definecolor{cyan}{rgb}{0,.5,.5}
\newcommand\ar[2]{\sout{#1} \textcolor{cyan}{#2}}
\newcommand\arc[1]{\textcolor{cyan}{#1}}
% light brown for KT
\definecolor{lightbrown}{rgb}{0.5,0.5,0}
\newcommand\kt[2]{\sout{#1} \textcolor{lightbrown}{#2}}
\newcommand\ktc[1]{\textcolor{lightbrown}{#1}}

@@

@@

\bibliographystyle{apalike}
\usepackage{natbib}
\usepackage{fullpage}

@@

@@

We do not see any immediate reason that this model could not be extended to cover a larger range of the data, and chose the value $t_0 = \text{2010-11-13}$.
This choice of $t_0$ corresponds to using approximately $1\%$ of the available data to determine initial values of the latent states.
In addition to modeling a larger portion of the available data, this choice of $t_0$ corresponds to an important real-world event, as daily reports from each of the departments were not being sent to the health ministry until November 10, 2010; % TODO: Cite
this choice of $t_0$ therefore makes $\bm{Y}\big(t_1\big)$ the first week of data once daily reports are being sent to the health ministry.
The few observation times that exist before $t_0$ are used to initialize the model by enforcing model dynamics on these preliminary observations.
For convenience, we denote these observations as $t_{-3}, t_{-2}$ and $t_{-1}$; as before, we let $y_{u,-k} = Y_{u}\big(t_{-k}\big)$ denote the observed case count for unit $u$ at time point $t_{-k}$, where $k \in 1:3$.
Equations for the initial values of non-zero latent states are provided in Eqs.~\myeqref{eq:model3:I0}--\myeqref{eq:model3:B0}; these equations match those that were used by \citet{lee20}, the primary change being a change to the value of $t_0$.

@@

% Argue why we didn't need to do this for Model 2.

% TODO: Check this section.

@@

mod3_cases <- as.data.frame(t(h3@data))
colnames(mod3_cases) <- deps
m3_inits <- c(7362, 1914, 22, 8, 2818, 82, 2499, 6953, 44, 7) # TODO: Also fit by hand, but could be changed to initializing to the observation before.
names(m3_inits) <- deps

@@

In this section, we provide more detail on the process that was used to estimate the coefficients used in Model~3.
In particular, we discuss why we decided to include additional model parameters---those that are associated with the behavior of the system during Hurricane Matthew---that were not considered by \cite{lee20}.
To calibrate this model, we used the iterated block particle filter (IBPF) method of \cite{ionides22}.
Due to the novelty of this algorithm, there does not exist published examples the IBPF algorithm used for data analysis outside of the papers in which the algorithm was introduced \citep{ning21ibpf,ionides22}, which at least partially motivates the inclusion additional details in this section.

\cite{lee20} were only able to estimate model parameters to a simplified version of Model~3 on a subset of the available data, as no method existed at the time of their publication to fit a fully coupled meta-population model to disease incidence data.
Building on their results, we fit the fully coupled version of Model~3 to nearly all available data, reserving only a few observations to use to calibrate the initial conditions of the model (see the previous section for more details).
In order to properly maximize the model likelihood, we found that it was necessary to use multiple searches for the MLE, periodically pruning away less successful searches.
The first collection of searches was performed by obtaining initial values for the parameters by uniformly sampling values from a predefined hypercube.
A subsequent refinement search used starting values corresponding to the largest model likelihoods obtained in the previous round.
This is similar to the approach used by \cite{ionides22}.

We use the technique described above to fit the fully coupled version of Model~3 proposed by \cite{lee20}.
The maximum likelihood we obtained with the model was $\Sexpr{round(max(haiti3_sub_fit$search2$logLiks$logLik))}$, which is higher than the benchmark model ($\Sexpr{round(sum(unitLikes_bench$ll))}$).
While beating a simple associative benchmark is promising, this does not immediately rule out the possibility that the model is a poor description of the system.
Additional investigation of parameters estimates and their corresponding implications on model based conclusions should always be conducted.
For meta-population models, it is worth considering how well the model fits the data to each spatial unit.
This can be done by looking at conditional log-likelihoods, which is part of the output of the \texttt{bpfilter} algorithm in the \texttt{spatPomp} package.
The likelihoods for each department, compared to the corresponding benchmark model, are displayed in Fig.~\ref{fig:h3UnitLikes}.
The figure demonstrates that while the log-likelihood of the fitted model outperforms the auto-regressive negative binomial benchmark model, Model~3 has lower likelihoods than the benchmark for some departments.

\begin{figure}[ht]
<<h3UnitLikes, fig.height=3, fig.width=5.5, echo=FALSE>>=
ggplot(bench_M3_joined, aes(x = ll, y = reorder(unit, ll))) +
  geom_point(aes(col = type)) +
  ylab("Unit") +
  xlab("Log Likelihood") +
  scale_color_manual(
    labels = c("ll" = "Benchmark", "ll_est" = "Model 3"),
    values = c("ll" = "#1b9e77", "ll_est" = "#d95f02")
  ) +
  theme(legend.title = element_blank(),
        axis.title.y = element_blank())
@
\caption{\label{fig:h3UnitLikes}Log-likelihoods of Model~3 for each department compared to the corresponding benchmark model prior to the inclusion of parameters for initial values and Hurricane Matthew.}
\end{figure}

>>>>

%% copied from pnas-new.cls
\usepackage{amsmath,amsfonts,amssymb,graphicx}
\usepackage{pict2e}
\usepackage{curve2e}

%%%% parameters %%%%%%%%%%5
\newcommand\Wsat{W_{\mathrm{sat}}}
\newcommand\muIR{\mu_{IR}}
\newcommand\muEI{\mu_{EI}}
\newcommand\transmission{\beta}
\newcommand\seasAmplitude{a}
\newcommand\rainfallExponent{r}
\newcommand\muRS{\mu_{RS}}
\newcommand\vaccineEfficacy{\theta}
\newcommand\muBirth{\mu_S}
\newcommand\muDeath{\delta}
\newcommand\choleraDeath{\delta_{C}}
\newcommand\symptomFrac{f}
\newcommand\asymptomRelativeInfect{\epsilon}
\newcommand\asymptomRelativeShed{\epsilon_{W}}
\newcommand\Wbeta[1]{\beta_{W#1}}
\newcommand\Wremoval{\delta_W}
\newcommand\Wshed{\mu_W}
\newcommand\mixExponent{\nu}
\newcommand\sigmaProc{\sigma_{\mathrm{proc}}}
\newcommand\reportRate{\rho}
\newcommand\obsOverdispersion{\psi}
\newcommand\phaseParm{\phi}
\newcommand\transmissionTrend{\zeta}
\newcommand\Binit{\xi}
\newcommand\vaccClass{Z}
\newcommand\vaccCounter{z}
\newcommand\modelCounter{m}
\newcommand\Whur[1]{\beta_{W#1}^{hm}}
\newcommand\hHur[1]{h_{#1}^{hm}}
\newcommand\Iinit{I_{0,0}}
\newcommand\Einit{E_{0, 0}}
\newcommand\reportChange{\delta}
\newcommand\NBintercept{\alpha}
\newcommand\NBsize{\varphi}

@@

%% EDITING MACROS
\usepackage{color}
\usepackage[section]{placeins}
\usepackage[normalem]{ulem}% to use \sout in feedback commands
\usepackage{bm}
% orange for EI
\definecolor{orange}{rgb}{1,0.5,0}
\newcommand\ei[2]{\sout{#1} \textcolor{orange}{#2}}
\newcommand\eic[1]{\textcolor{orange}{[#1]}}
% green for JW
\definecolor{green}{rgb}{0,0.5,0}
\newcommand\jw[2]{\sout{#1} \textcolor{green}{#2}}
\newcommand\jwc[1]{\textcolor{green}{[#1]}}
% purple for JJ
\definecolor{purple}{rgb}{0.5,0,1}
\newcommand\jj[2]{\sout{#1} \textcolor{purple}{#2}}
\newcommand\jjc[1]{\textcolor{purple}{[#1]}}
% cyan for AR
\definecolor{cyan}{rgb}{0,.5,.5}
\newcommand\ar[2]{\sout{#1} \textcolor{cyan}{#2}}
% \newcommand\arc[1]{\textcolor{cyan}{#1}}
% light brown for KT
\definecolor{lightbrown}{rgb}{0.5,0.5,0}
\newcommand\kt[2]{\sout{#1} \textcolor{lightbrown}{#2}}
\newcommand\ktc[1]{\textcolor{lightbrown}{#1}}

@@

\makeatletter
\AtBeginDocument{%
  \expandafter\renewcommand\expandafter\subsection\expandafter{%
    \expandafter\@fb@secFB\subsection
  }%
}
\makeatother

@@

\bibliographystyle{apalike}
\usepackage{natbib}
\usepackage{fullpage}
% \usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
% \usepackage[labelfont=bf,labelsep=period]{caption} % For some reason this doesn't work with .Rnw file, but it does with just the tex file, so after compiling you can edit this so that it works.
\renewcommand{\figurename}{Fig}

@@

\section{Model Diagrams}

Each of the dynamic models considered in this manuscript can be fully described using the model descriptions in the manuscript, coupled with the additional information described in Sections 2 and 3 of this supplement.
Despite this, diagrams of dynamic systems are often helpful to understand the equations.
In this section, we give three diagrams representing Models~1--3, respectively.
Because the models are defined by their mathematical equations and numeric implementation, these diagrams are not unique visual representations of the model.
Alternative representations that may be helpful in understanding the models explored in this paper were made by \citet{lee20sup}.

\begin{figure}[!h]
%%%%% SEAIR diagram
\begin{center}
  \resizebox{0.9\textwidth}{!}{
    \Large
    \setlength{\unitlength}{5pt}
    \begin{picture}(100,95)(0,15)
      \thicklines

      % Unvaccinated Compartments
      \put(10,63){\framebox(6,6){$\mathrm{S}_0$}}
      \put(34,63){\framebox(6,6){$\mathrm{E}_0$}}
      \put(58,71){\framebox(6,6){$\mathrm{A}_0$}}
      \put(82,63){\framebox(6,6){$\mathrm{R}_0$}}
      \put(58,55){\framebox(6,6){$\mathrm{I}_0$}}

      % Births
      \put(1, 66){\vector(1, 0){8}}
      \put(4, 68){$\muBirth$}

      % S_0 -> E_0
      \put(17,66){\vector(1,0){16}}
      \put(22,67){$\lambda(t)$}

      % E_0 -> A_0
      \put(41,67.5){\vector(2.1,1){16}}
      % \put(37, 72){$\muEI\big(1-\symptomFrac_{\vaccCounter}(t)\big)$}

      % E_0 -> I_0
      \put(41,65.5){\vector(2.1,-1){16}}

      % A_0 -> R_0
      \put(65,74){\vector(2.1,-1){16}}

      % I_0 -> R_0
      \put(65,59){\vector(2.4,1){16}}

      % R_0 -> S_0
      \cbezier(84,70)(75, 95)(23,95)(14,70)
      \put(14,70){\vector(-1, -2){0.3}}

      % Unvaccinated death rates
      \put(13, 70){\vector(0, 1){10}}
      \put(11, 74){$\muDeath$}

      \put(37, 70){\vector(0, 1){10}}
      \put(35, 74){$\muDeath$}

      \put(61, 78){\vector(0, 1){9}}
      \put(59, 80){$\muDeath$}

      \put(61, 62){\vector(1.5, 2){6}}
      \put(61.5, 65.5){$\muDeath$}

      \put(85, 70){\vector(0, 1){10}}
      \put(83, 74){$\muDeath$}

      % Vaccinated Compartments
      \put(10,37){\framebox(6,6){$\mathrm{S}_\vaccCounter$}}
      \put(34,37){\framebox(6,6){$\mathrm{E}_\vaccCounter$}}
      \put(58,45){\framebox(6,6){$\mathrm{A}_\vaccCounter$}}
      \put(82,37){\framebox(6,6){$\mathrm{R}_\vaccCounter$}}
      \put(58,29){\framebox(6,6){$\mathrm{I}_\vaccCounter$}}

      % S -> E
      \put(17,40){\vector(1,0){16}}
      \put(22,41){$\lambda(t)$}

      % E -> A
      \put(41,40.5){\vector(2.1,1){16}}

      % E -> I
      \put(41,39.5){\vector(2.1,-1){16}}

      % A -> R
      \put(65,48){\vector(2.1,-1){16}}

      % I -> R
      \put(65,32){\vector(2.1,1){16}}

      % R -> S
      \cbezier(84, 36)(75, 10)(23, 10)(14, 36)
      \put(14,36){\vector(-1, 2){0.3}}

      % Vaccinated death rates
      \put(13, 36){\vector(0, -1){10}}
      \put(11, 31){$\muDeath$}

      \put(37, 36){\vector(0, -1){10}}
      \put(35, 31){$\muDeath$}

      \put(61, 28.5){\vector(0, -1){9}}
      \put(59, 24){$\muDeath$}

      \put(61, 44){\vector(1.5, -2){6}}
      \put(62, 38){$\muDeath$}

      \put(85, 36){\vector(0, -1){10}}
      \put(83.2, 30.2){$\muDeath$}

      % S_0 -> S_\vaccCounter
      \multiput(13,62)(0,-2){9}{\line(0,-1){1}}
      \put(13, 43.5){\vector(0, -1){0}}

      % E_0 -> E_\vaccCounter
      \multiput(37,62)(0,-2){9}{\line(0,-1){1}}
      \put(37, 43.5){\vector(0, -1){0}}

      % R_0 -> R_\vaccCounter
      \multiput(85,62)(0,-2){9}{\line(0,-1){1}}
      \put(85, 43.5){\vector(0, -1){0}}

      % A_0 -> A_\vaccCounter
      \multiput(57.5,72)(-1.9, 0){3}{\line(-1, 0){0.9}}
      \put(52.2, 72){\line(-2.1, -1){1}}
      \multiput(51,71)(0, -2){11}{\line(0, -1){1}}
      \put(52.2, 49){\line(-2.1, 1){1}}
      \multiput(54,49)(1.9, 0){2}{\line(-1, 0){0.9}}
      \put(57, 49){\vector(1, 0){0}}

      % I_0 -> I_\vaccCounter
      \multiput(65, 57.5)(1.9, 0){3}{\line(1, 0){0.9}}
      \put(70.6, 57.5){\line(2.1, -1){1}}
      \multiput(72, 56.5)(0, -2){13}{\line(0, -1){1}}
      \put(72, 30.9){\line(-2, -1.2){0.9}}
      \multiput(65.6, 30.2)(1.9, 0){3}{\line(1, 0){0.9}}
      \put(64.5, 30.2){\vector(-1, 0){0}}

    \end{picture}
  }
\end{center}
\caption{A flow diagram for the SEAIR model.}\label{fig:flow_diagram}
\end{figure}

\begin{figure}[!h]
%%%%% SEAIR diagram
\begin{center}
  \resizebox{\textwidth}{!}{
    \Large
    \setlength{\unitlength}{5pt}
    \begin{picture}(100,95)(0,15)
      \thicklines

      % Unvaccinated Compartments
      \put(10,63){\framebox(6,6){$\mathrm{S}_{u0}$}}
      \put(34,69){\framebox(6,6){$\mathrm{E}_{u0}$}}
      \put(58,77){\framebox(6,6){$\mathrm{A}_{u0}$}}
      \put(82,77){\framebox(6,6){$\mathrm{R^A}_{u0}$}}
      \put(82,61){\framebox(6,6){$\mathrm{R^I}_{u0}$}}
      \put(58,61){\framebox(6,6){$\mathrm{I}_{u0}$}}

      % % S_0 -> E_0
      \put(17,66){\vector(3,1.1){16}}
      % \put(22,67){$\lambda(t)$}

      % % E_0 -> A_0
      \put(41,72.5){\vector(2.1,1){16}}
      % % \put(37, 72){$\muEI\big(1-\symptomFrac_{\vaccCounter}(t)\big)$}

      % % E_0 -> I_0
      \put(41,71.5){\vector(2.1,-1){16}}

      % % A_0 -> R_0
      \put(65,80.5){\vector(1,0){16}}

      % A_0 -> W_u
      % \multiput(65, 79.8)(1, -1){8}{\line(1, -1){0.25}}

      % I_0 -> W_u
      % \multiput(65, 64.2)(1, 1){8}{\line(1, 1){0.25}}

      % Long verticle dotted line
      % \multiput(72.9, 72)(0, -1.41){13}{\line(0, -1){0.25}}

      % Long backwards horizontal dotted line
      % \multiput(73, 54)(-1.41, 0){22}{\line(0, -1){0.25}}
      % \multiput(73, 34)(0, 1.41){15}{\line(0, 1){0.25}}
      % \put(42, 53.9){\vector(-1, 0){0.2}}

      % A_uz -> W_u
      \put(58,39){\framebox(6,6){$\mathrm{A}_{u\vaccCounter}$}}
      \qbezier[25](60.5, 46)(51, 52)(42, 53)
      \put(42, 53){\vector(-1, 0.1){0.5}}

      % A_uz -> W_u
      \qbezier[25](60.5, 60)(51, 54)(42, 54)
      \put(42, 54){\vector(-1, -0.1){0.5}}

      % I_uz -> W_u
      \qbezier[30](60.5,30)(48, 45)(40, 50)
      \put(40, 50){\vector(-1, 0.8){0.5}}

      % A_u0 -> W_u
      \qbezier[29](60.5,76)(48, 61)(40.6, 57)
      \put(40.8, 57.1){\vector(-1, -0.7){0.5}}

      % A + I_\vaccCounter -> W_u
      % \multiput(65, 41.8)(1, -1){8}{\line(1, -1){0.25}}
      % \multiput(65, 26.2)(1, 1){8}{\line(1, 1){0.25}}

      % W_u -> backwards diags.
      % \multiput(33, 57)(-1, 1.2){10}{\line(-1, 1.2){0.1}}
      % \put(24.5, 67){\vector(-1, 1.2){0.5}}
      % \multiput(33, 51)(-1, -1.2){10}{\line(-1, -1.2){0.1}}
      % \put(24, 40){\vector(-1, -1.2){0.5}}
      %\put(65, 80){\vector(2, -1){8}}

      % % I_0 -> R_0
      \put(65,63.5){\vector(1,0){16}}

      % % Vaccinated Compartments
      \put(10,37){\framebox(6,6){$\mathrm{S}_{u\vaccCounter}$}}
      \put(34,31){\framebox(6,6){$\mathrm{E}_{u\vaccCounter}$}}
      \put(58,39){\framebox(6,6){$\mathrm{A}_{u\vaccCounter}$}}
      \put(82,39){\framebox(6,6){$\mathrm{R^A}_{u\vaccCounter}$}}
      \put(58,23){\framebox(6,6){$\mathrm{I}_{u\vaccCounter}$}}
      \put(82,23){\framebox(6,6){$\mathrm{R^I}_{u\vaccCounter}$}}

      % R_0 -> S_0
      \cbezier(85,22)(80, 8)(18,8)(12,36)
      \put(12.1,35.5){\vector(-1, 3.5){0.3}}

      \cbezier(85,38)(80, 12)(18,15)(14,36)
      \put(14.1,35.5){\vector(-1, 3.5){0.3}}

      % R_0 -> S_0
      \cbezier(85,84)(80, 100)(18,100)(12,70)
      \put(12.1,70.5){\vector(-1, -3.5){0.3}}

      \cbezier(85,68)(80, 95)(18, 92)(14,70)
      \put(14.1,70.5){\vector(-1, -3.5){0.3}}

      % % S -> E
      \put(17,40){\vector(3,-1.1){16}}
      % \put(22,41){$\lambda(t)$}

      % % E -> A
      \put(41,34.5){\vector(2.1,1){16}}

      % % E -> I
      \put(41,33.5){\vector(2.1,-1){16}}

      % % A -> R
      \put(65,42.5){\vector(1,0){16}}

      % % I -> R
      \put(65,25.5){\vector(1,0){16}}

      % S_0 -> S_\vaccCounter
      \multiput(12,62)(0,-2){9}{\line(0,-1){1}}
      \put(12, 44){\vector(0, -1){0}}

      % S_0 -> S_\vaccCounter
      % Wanning vaccine efficacy
      \multiput(14,44)(0,2){9}{\line(0,1){1}}
      \put(14, 62){\vector(0, 1){0}}

      \put(37, 54){\circle{7}}
      \put(35, 53){$\mathrm{W_u}$}

      \qbezier[10](32.5, 54)(29.5, 54)(26.5, 54)
      \put(26, 54){\vector(-1, 0){0.1}}
      \put(28, 55){$\Wremoval$}

    \end{picture}
  }
\end{center}
\caption{A flow diagram for the SEAIR model 2. This is a constant population model, there are no births/deaths. Vaccinations are assumed to only be given to susceptible individuals, and vaccine immunity wanes only with susceptible vaccinated individuals.}\label{fig:flow_diagram2}
\end{figure}

\begin{figure}[!h]
%%%%% SEAIR diagram
\begin{center}
  \resizebox{\textwidth}{!}{
    \Large
    \setlength{\unitlength}{5pt}
    \begin{picture}(100,85)(0,15)

    % COMPARTMENTS
    % \put(36, 47){\framebox(6, 6){$\mathrm{W}_{u}$}}

    \put(39, 50){\circle{6}}
    \put(37, 49){$\mathrm{W}_{u}$}

    \put(8, 55){\framebox(6, 6){$\mathrm{S_{u0}}$}}
    \put(8, 39){\framebox(6, 6){$\mathrm{S_{u\vaccCounter}}$}}
    % \put(22, 45){\circle{2}}
    % \put(22, 55){\circle{2}}
    \put(36, 62.5){\framebox(6, 6){$\mathrm{A_{u}}$}}
    \put(36, 31.5){\framebox(6, 6){$\mathrm{I_{u}}$}}
    \put(56, 55){\framebox(6, 6){$\mathrm{R^1_{u0}}$}}
    \put(71, 55){\framebox(6, 6){$\mathrm{R^2_{u0}}$}}
    \put(86, 55){\framebox(6, 6){$\mathrm{R^3_{u0}}$}}
    \put(56, 39){\framebox(6, 6){$\mathrm{R^1_{u\vaccCounter}}$}}
    \put(71, 39){\framebox(6, 6){$\mathrm{R^2_{u\vaccCounter}}$}}
    \put(86, 39){\framebox(6, 6){$\mathrm{R^3_{u\vaccCounter}}$}}
    % \dottedline{3}(0,5)(70,5)

    % INDIVIDUAL MOVEMENT
    % \put(15, 58){\line(2.4, -1){5.5}}
    % \put(15, 42){\line(2.4, 1){5.5}}

    % \put(23, 46.3){\vector(1, 1.45){12}}
    % \put(23, 53.7){\vector(1, -1.45){12}}

    % \put(23.5, 55.7){\vector(1.1, 1){11}}
    % \put(23.5, 44.3){\vector(1.1, -1){11}}

    % Backward Diags, W_u
    % \multiput(35, 51)(-1, .3){12}{\line(-1, .3){0.2}}
    % \put(24, 54.4){\vector(-1, 0.3){0.5}}
    % \multiput(35, 49)(-1, -0.3){12}{\line(-1, -0.3){0.15}}
    % \put(24, 45.6){\vector(-1, -0.3){0.5}}

    % S_u0 -> A_u
    \put(15, 58.5){\vector(3, 1.2){20}}

    % S_u0 -> I_u
    \put(15, 57.5){\vector(1, -1.1){20}}

    % S_uz -> I_u
    \put(15, 41.5){\vector(3, -1.2){20}}

    % S_uz -> A_u
    \put(15, 42.5){\vector(1, 1.1){20}}

    % I_u -> R_u0
    \put(43, 66){\vector(2, -1){12}}

    % A_u -> R_uz
    \multiput(43, 64)(1.25, -1.6){12}{\line(1, -1.3){0.7}}
    \put(57, 46){\vector(1, -1.3){0.2}}

    \put(43, 36){\vector(1.7, 2){15}}

    \multiput(60, 54)(0, -2){4}{\line(0, -1){1}}
    \put(60, 46){\vector(0, -1){0.5}}

    \multiput(74, 54)(0, -2){4}{\line(0, -1){1}}
    \put(74, 46){\vector(0, -1){0.5}}

    \multiput(89, 54)(0, -2){4}{\line(0, -1){1}}
    \put(89, 46){\vector(0, -1){0.5}}

    \put(63, 58){\vector(1, 0){7}}
    \put(78, 58){\vector(1, 0){7}}

    \put(63, 42){\vector(1, 0){7}}
    \put(78, 42){\vector(1, 0){7}}

    \multiput(39, 61.5)(0, -1){8}{\line(0, -1){0.3}}
    \put(39,54){\vector(0, -1){0.5}}

    \multiput(39, 38.5)(0, 1){8}{\line(0, 1){0.3}}
    \put(39,46){\vector(0, 1){0.5}}

    \cbezier(89,62)(80, 85)(20, 85)(12,62)
    \put(12.2,62.5){\vector(-1, -3.5){0.3}}

    \cbezier(89,38)(80, 15)(20, 15)(12,38)
    \put(12.2,37.5){\vector(-1, 3.5){0.3}}

    \multiput(11, 54)(0, -2){4}{\line(0, -1){1}}
    \put(11, 46){\vector(0, -1){0.5}}

    % DEMOGRAPHY
    % -> S
    \put(0, 58){\vector(1, 0){7}}
    \put(1, 59){$\mathrm{\mu_{\demography S_{u0}}}$}

    % S Death
    \put(10, 62){\vector(0, 1){7}}
    \put(8, 64){$\mathrm{\muDeath}$}
    \put(10, 38){\vector(0, -1){7}}
    \put(8, 34){$\mathrm{\muDeath}$}

    % A + I death
    \put(39, 69.5){\vector(0, 1){7}}
    \put(37, 71.5){$\mathrm{\muDeath}$}

    \put(39, 31.5){\vector(0, -1){7}}
    \put(30, 27.5){$\mathrm{\muDeath + \choleraDeath}$}

    % R Death
    \put(59, 62){\vector(0, 1){7}}
    \put(57, 64){$\mathrm{\muDeath}$}

    \put(74, 62){\vector(0, 1){7}}
    \put(72, 64){$\mathrm{\muDeath}$}

    \put(59, 38){\vector(0, -1){7}}
    \put(57, 34){$\mathrm{\muDeath}$}

    \put(74, 38){\vector(0, -1){7}}
    \put(72, 34){$\mathrm{\muDeath}$}

    \put(93, 42){\vector(1, 0){7}}
    \put(95.5, 43){$\mathrm{\muDeath}$}

    \put(93, 58){\vector(1, 0){7}}
    \put(95.5, 59){$\mathrm{\muDeath}$}

    % W death
    % \put(43, 50){\vector(1, 0){7}}
    \multiput(43, 50)(1, 0){7}{\line(1, 0){0.2}}
    \put(50.2, 50){\vector(1, 0){0.5}}
    \put(44, 51){$\mathrm{\Wremoval}$}
    % \qbezier[12](10, 20)(20, 30)(25, 20)

    \end{picture}
  }
\end{center}
\caption{A flow diagram for the SAIR model 3. This model assumes a constant population while also including a mechanism for births/deaths; all deaths are balanced by births into the unvaccinated susceptible compartment, so the birth rate $\mu_{\demography S_{u0}}$ corresponds to the sum total deaths from the remaining compartments. The model assumes that symptomatic individuals will not be vaccinated, hence no vaccination arrow exiting the $I_u0$ compartment.}\label{fig:flow_diagram3}
\end{figure}

@@

We do not see any immediate reason that this model could not be extended to cover a larger range of the data, and chose the value $t_0 = \text{2010-11-13}$.
This choice of $t_0$ corresponds to using approximately $1\%$ of the available data to determine initial values of the latent states.
In addition to modeling a larger portion of the available data, this choice of $t_0$ corresponds to an important real-world event, as daily reports from each of the departments were not being sent to the health ministry until November 10, 2010 \citep{barzilay13};
this choice of $t_0$ therefore makes $\bm{Y}\big(t_1\big)$ the first week of data once daily reports are being sent to the health ministry.
The few observation times that exist before $t_0$ are used to initialize the model by enforcing model dynamics on these preliminary observations.
For convenience, we denote these observations as $t_{-3}, t_{-2}$ and $t_{-1}$; as before, we let $y_{u,-k} = Y_{u}\big(t_{-k}\big)$ denote the observed case count for unit $u$ at time point $t_{-k}$, where $k \in 1:3$.
Equations for the initial values of non-zero latent states are provided in Eqs.~\myeqref{eq:model3:I0}--\myeqref{eq:model3:B0}; these equations match those that were used by \citet{lee20}, the primary change being a change to the value of $t_0$.

@@

@@

mod3_cases <- as.data.frame(t(h3@data))
colnames(mod3_cases) <- deps
m3_inits <- c(7362, 1914, 22, 8, 2818, 82, 2499, 6953, 44, 7)  # fit by hand
names(m3_inits) <- deps

@@

In this section, we provide more detail on the process that was used to estimate the coefficients of Model~3.
In particular, we discuss why we decided to include additional model parameters---those that are associated with the behavior of the system during Hurricane Matthew---that were not considered by \cite{lee20}.
To calibrate this model, we used the iterated block particle filter (IBPF) method of \cite{ionides22}.
Due to the novelty of this algorithm, there does not exist published examples the IBPF algorithm used for data analysis outside of the papers in which the algorithm was introduced \citep{ning21ibpf,ionides22}, which is one motivation of the inclusion additional details related to fitting and diagnosing the model fit provided here.

\cite{lee20} were only able to estimate model parameters to a simplified version of Model~3 on a subset of the available data, as no method existed at the time of their publication to fit a fully coupled meta-population model to disease incidence data.
Building on their results, we fit the fully coupled version of Model~3 to nearly all available data, reserving only a few observations to use to calibrate the initial conditions of the model (see the previous section for more details).
In order to properly maximize the model likelihood, we found that it was necessary to use multiple searches for the MLE, periodically pruning away less successful searches.
The first collection of searches was performed by obtaining initial values for the parameters by uniformly sampling values from a predefined hypercube.
A subsequent refinement search used parameter values corresponding to the largest model likelihoods as starting parameter values.
The need for multiple searches does not appear to be uncommon, as a similar approach was used by \cite{ionides22}.

We use the technique described above to fit the fully coupled version of Model~3 proposed by \cite{lee20}.
The maximum likelihood we obtained after two rounds of searching was $\Sexpr{round(max(haiti3_sub_fit$search2$logLiks$logLik))}$, which is higher than the benchmark model ($\Sexpr{round(sum(unitLikes_bench$ll))}$).
While beating a simple associative benchmark is promising, this does not immediately rule out the possibility that the model is a poor description of the system.
Additional investigation of parameters estimates and their corresponding implications on model based conclusions should always be conducted.
For meta-population models, it is worth considering how well the model fits the data to each spatial unit.
% This can be done by looking at conditional log-likelihoods, which is part of the output of the \texttt{bpfilter} algorithm in the \texttt{spatPomp} package.
The likelihoods for each department, compared to the corresponding benchmark model, are displayed in Fig.~\ref{fig:h3UnitLikes}.
The figure demonstrates that while the log-likelihood of the fitted model outperforms the auto-regressive negative binomial benchmark model, Model~3 has lower likelihoods than the benchmark for some departments.

\begin{figure}[!h]
<<h3UnitLikes, fig.height=3, fig.width=5.5, echo=FALSE>>=
ggplot(bench_M3_joined, aes(x = ll, y = reorder(unit, ll))) +
  geom_point(aes(col = type)) +
  ylab("Unit") +
  xlab("Log Likelihood") +
  scale_color_manual(
    labels = c("ll" = "Benchmark", "ll_est" = "Model 3"),
    values = c("ll" = "#1b9e77", "ll_est" = "#d95f02")
  ) +
  theme(legend.title = element_blank(),
        axis.title.y = element_blank())
@
\caption{\label{fig:h3UnitLikes}Log-likelihoods of Model~3 for each department compared to the corresponding benchmark model prior to the inclusion of parameters that account for Hurricane Matthew.}
\end{figure}

In addition to considering the conditional log-likelihoods of each unit, one can consider conditional log-likelihoods of each observation.
When compared to a benchmark, this level of detail can provide useful information about which observations are well described by the model and which are not.
In Fig.~\ref{fig:condLL}, we plot the conditional log-likelihoods of Model~3 for each observation.
Typically it is most useful to compare the conditional log-likelihoods of the model under consideration to a benchmark, as plotting only conditional log-likelihoods without a comparison may not be helpful.
In this case, however, the same insight can be drawn using a figure without a benchmark comparison, so we do not include the benchmark in order to avoid the issue of over-plotting.

\begin{figure}[!h]
<<CondLLFig, echo=FALSE, message=FALSE, fig.height=3>>=
